\section{Advanced Topics in Computability Theory}

\subsection{A Definition of Information}

We define the quantity of information contained in an object to be the size of that object's smallest representation or description.

\subsubsection{Minimal Length Descriptions}

We describe a binary string $x$ with a Turing machine $M$ and a binary input $w$ to $M$. The length of the description is the combined  length of representing $M$ and $w$. 

We define the string $\desc{M, w}$ to be $\desc{M}w$, where we simply concatenate the binary string $w$ onto the end of the binary encoding of $M$.

\begin{shaded}
\textbf{ITOC - Definition 6.23}

\medskip
Let $x$ be a binary string. The \textbf{minimal description} of $x$, written $d(x)$, is the shorted string $\desc{M, w}$ where TM $M$ on input $w$ halts with $x$ on its tape. The \textbf{descriptive complexity} of $x$, written $K(x)$, is $K(x) = |d(x)|$.
\end{shaded}

{\color{blue} $K(x)$ is the length of the minimal description of $x$}.


\begin{shaded}
\textbf{ITOC - Theorem 6.24}
\[
\exists c \; \forall x \;  [K(x) \leq |x| + c]
\]
\end{shaded}

The descriptive complexity of a string is at most a fixed constant more than its length. The constant is a universal one, not dependent on the string.

\begin{mdframed}
\begin{proof}
Let $M$ be a Turing machine that halts as soon as it is started. A description of $x$ is simply $\desc{M}x$, $c$ is the length of $\desc{M}$.
\end{proof}
\end{mdframed}

The amount of information contained by a string cannot be (substantially) more than its length.

\begin{shaded}
\textbf{ITOC - Theorem 6.25}
\[
\exists c \; \forall x \;  [K(xx) \leq K(x) + c]
\]
\end{shaded}

\begin{mdframed}
\begin{proof}
Consider TM $M$, which expects an input of the form $\desc{N, w}$, where $N$ is a TM and $w$ is an input for it.

$M$ = On input $\desc{N, w}$
\begin{enumerate}
\item Run $N$ on $w$ until it halts and produces an output string $s$.
\item Output string ss.
\end{enumerate}

A description of $xx$ is $\desc{M}d(x)$. The length of this description is $|\desc{M}| + |d(x)| = c + K(x)$, where $c$ is the length of $\desc{M}$.
\end{proof}
\end{mdframed}

\begin{shaded}
\textbf{ITOC - Theorem 6.25}
\[
\exists c \; \forall x,y \;  [K(xy) \leq 2K(x) _ K(y) + c]
\]

can be improved to
\[
K(xy) \leq 2\log_2(K(x)) + K(x) + K(y) + c
\]
\end{shaded}

\begin{shaded}
For any description language $p$, a fixed constant $c$ exists that depends only on $p$ where
\[
\forall x \; [K(x) \leq K_p(x) + c]
\]
\end{shaded}

\subsubsection{Incompressible Strings and Randomness}

\begin{shaded}
\textbf{ITOC - Definition 6.28}

\medskip
Let $x$ be a string. Say that $x$ is \textbf{$c$-compressible} if
\[
K(x) \leq |x| - c
\]
If $x$ is not $c$-compressible, say that $x$ is \textbf{incompressible  by $c$}.

If $x$ is incompressible by 1, say that $x$ is \textbf{incompressible}.
\end{shaded}

\begin{shaded}
\textbf{ITOC - Theorem 6.29}

\medskip
Incompressible strings of every length exist.
\end{shaded}